---
title: "Mean-Block models: clustering variables by their response to covariates"
author: "Julien Chiquet & Jeanne Tous"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
bibliography: references.bib
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Mean-Block models: clustering variables by their response to covariates}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Preliminaries

This vignette introduces the **Mean-Block** model (`model = "mean"`), the second
model family shipped with `normalblockr`. Where the Normal-Block model of the
`normal-block` vignette puts the clustering in the *latent covariance*, this one
puts it in the *mean*: variables are grouped by how their expected value responds
to the covariates.

We use the same dataset as the `breast-cancer-proteomics` vignette (`?brca_rppa`)
on purpose, so that the two families can be read side by side on data you have
already seen: 163 proteins measured on 346 breast-cancer tumor samples, with each
sample's PAM50 molecular subtype as the covariate.

```{r, set-up}
library(normalblockr)
```

# Mathematical background

Write $Z \in \mathbb{R}^{n \times p}$ for the (log-)responses and
$X \in \mathbb{R}^{n \times d}$ for the design matrix. Each observation is
Gaussian, with a mean built from a *clustering of the variables*:

$$Z_i \sim \mathcal{N}(\mu_i, \Sigma), \qquad \mu_i = C B^\top X_i$$

where

- $B \in \mathbb{R}^{d \times q}$ holds one regression profile **per cluster**,
  so $B^\top X_i \in \mathbb{R}^q$ is the linear predictor of each of the $q$
  clusters for observation $i$;
- $C \in \{0,1\}^{p \times q}$ assigns every variable to exactly one cluster, and
  maps those $q$ cluster-level predictors back onto the $p$ variables;
- $\Sigma \in \mathbb{R}^{p \times p}$ is the residual covariance between
  variables.

The modelling assumption is that **all variables in a cluster share the same
regression profile**. This is what makes the model parsimonious: the regression
part costs $d \times q$ parameters instead of the $d \times p$ of an
unconstrained multivariate regression, with $q \ll p$.

Contrast this with the Normal-Block model of the other vignettes, where $C$
structures $\mathrm{Var}(Z_i)$ and the covariates only enter through a
variable-wise $B^\top X_i$: there, two variables are in the same cluster when
they *covary* the same way; here, when they *respond* the same way. The two
answer different questions and generally return different groupings.

### Known versus unknown clustering

When $C$ is known, $(B, \Sigma)$ are obtained by a plain EM alternating

$$\widehat B = (X^\top X)^{-1}(X^\top Z)\,\Sigma^{-1} C \,(C^\top \Sigma^{-1} C)^{-1},
\qquad \widehat\Sigma = \tfrac{1}{n} \widehat R^\top \widehat R,$$

with $\widehat R = Z - X \widehat B C^\top$ the residuals. One recognizes a
generalized-least-squares estimator: the unconstrained fit
$(X^\top X)^{-1}X^\top Z$ projected onto the space of coefficients compatible
with the clustering.

When $C$ is unknown it becomes a latent variable, with
$C_{j\bullet} \sim \mathcal{M}(1, \alpha)$. The observed likelihood then sums
over $q^p$ configurations and is intractable, so the fit maximizes a variational
lower bound (ELBO) under a mean-field approximation
$Q(C) = \prod_j Q_j(C_{j\bullet})$, parametrized by
$\tau \in [0,1]^{p \times q}$, $\tau_{jk} = \mathbb{E}_Q[C_{jk}]$:

$$J(\theta) = -\frac{np}{2}\log(2\pi) + \frac{n}{2}\log|\Sigma^{-1}|
 + \sum_{j,k} \tau_{jk}\log\alpha_k - \sum_{j,k}\tau_{jk}\log\tau_{jk}
 - \frac{1}{2}\Big[\mathrm{Tr}(R\Sigma^{-1}R^\top) + \mathrm{Tr}(\Phi B^\top X^\top X B)\Big]$$

with $R = Z - XB\tau^\top$ and $\Phi$ a correction term accounting for the
uncertainty on the assignments (it vanishes as $\tau$ becomes binary). The
variational E-step updates $\tau$ one variable at a time: with the other rows
held fixed the ELBO is linear in $\tau_{j\bullet}$ up to its entropy, so each
row's softmax is an *exact* maximization and a sweep can never decrease the
ELBO.

See @tous2026 for the Normal-Block model itself; the mean-block derivations
follow the same variational recipe.

# The data

```{r, data-load}
data(brca_rppa)
Y <- as.matrix(brca_rppa$expr)
X <- model.matrix(~ 0 + PAM50_SUBTYPE, data = brca_rppa$covariates)
nb_data <- NormalBlockData$new(Y, X)
dim(Y)
table(brca_rppa$covariates$PAM50_SUBTYPE)
```

`X` has one indicator column per PAM50 subtype, so each cluster's profile
$B_{\bullet k} \in \mathbb{R}^5$ is simply its mean expression level in each of
the five subtypes. Clustering the proteins therefore amounts to grouping them by
*subtype signature*.

# A known clustering

The dataset ships a Gene Ontology annotation (one biological-process term per
protein), which gives a clustering built with no reference to the model at all.
Handing it to `normal_block()` as a matrix fixes $C$: only $(B, \Sigma)$ are
estimated.

```{r, known-clustering}
go_term  <- factor(brca_rppa$gene_annotation$go_bp_term)
C_go     <- model.matrix(~ 0 + go_term)
NB_go    <- normal_block(nb_data, blocks = C_go, model = "mean",
                         control = NB_control(verbose = FALSE))
NB_go
```

The fitted `B` is a $5 \times q$ matrix of subtype profiles, one column per GO
term, and `fitted()` maps them back onto the proteins:

```{r, known-clustering-fitted, fig.width=6, fig.height=4}
dim(coef(NB_go))
plot(Y, fitted(NB_go), pch = ".", xlab = "observed", ylab = "fitted")
abline(0, 1, col = "red")
```

# Letting the model infer the clustering

## A collection over the number of clusters

With `blocks` a range of values, `normal_block()` returns one fitted model per
$q$. The initial clustering of each is derived from every protein's own fitted
profile $X\widehat B_{\bullet j}$ (see `NB_control(clustering_init = )`; `kmeans`
is this family's default).

```{r, collection}
NB_means <- normal_block(nb_data, blocks = seq(5, 120, by = 5), model = "mean",
                         control = NB_control(verbose = FALSE))
```

The grid is deliberately wide and coarse. With the default diagonal $\Sigma$
an extra cluster costs only $d = 5$ parameters, so the criteria stay hungry
for a long time: on a narrow range such as `1:15` they would still be
decreasing at the upper end, and "selecting" its boundary would mean nothing.

```{r, collection-criteria, fig.width=7, fig.height=4}
NB_means$plot(c("deviance", "BIC"))
```

```{r, collection-selection}
selected <- NB_means$get_best_model("ICL")
paste0("ICL selects ", selected$q, " clusters.")
```

That answer deserves a comment: the model groups 163 proteins into a few dozen
clusters, i.e. only a handful of proteins per cluster. It is saying that
subtype signatures are largely protein-specific here, with limited sharing --
a substantive finding about this dataset, not a failure of the fit. The
criteria do turn: they reach an interior minimum and rise again afterwards,
which is what makes the selection meaningful.

`refine()` is available to polish a collection, trying for each $q$ a short
split-and-reoptimize seeded from its $q-1$ neighbour and a merge from its
$q+1$ one, keeping a candidate only if it strictly lowers the deviance. It is
most useful on a contiguous range; on the coarse grid above there are no
adjacent $q$ to seed from, so it is skipped here.

## Reading the clusters

```{r, clusters}
table(selected$clustering)
```

Each cluster's profile across the five subtypes is a column of `coef()`; a
heatmap of that matrix is the most direct summary of what the model found.

```{r, cluster-profiles, fig.width=7, fig.height=4}
profiles <- coef(selected)
rownames(profiles) <- levels(brca_rppa$covariates$PAM50_SUBTYPE)
colnames(profiles) <- paste0("cluster ", seq_len(ncol(profiles)))
image(seq_len(nrow(profiles)), seq_len(ncol(profiles)), profiles,
      axes = FALSE, xlab = "", ylab = "", col = hcl.colors(20, "RdBu", rev = TRUE))
axis(1, seq_len(nrow(profiles)), rownames(profiles), las = 2, cex.axis = .7)
axis(2, seq_len(ncol(profiles)), colnames(profiles), las = 2, cex.axis = .7)
```

# Choosing the shape of the residual covariance

Everything above used the default residual covariance, `"diagonal"` (one
variance per variable). Two other shapes are available through
`NB_control(noise_covariance = )`: `"spherical"` (a single variance) and
`"full"` (the unconstrained $p \times p$ matrix). Only the last one has to be
inverted, so it alone requires $n > p$.

The default is deliberate. A full $\Sigma$ costs $p(p+1)/2 \approx 13{,}000$
parameters here, which drown the handful of mean parameters that BIC and ICL
are trying to weigh -- in a simulation study over 12 replicates, selecting $q$
by BIC was correct 10/12 times with a diagonal $\Sigma$ against 6/12 with a
full one at $n/p \approx 1.3$, *even when the data were generated with a full
$\Sigma$*. The quality of the clustering at a fixed $q$ was the same either
way; it is the choice of $q$ that suffers.

```{r, covariance-shapes}
shapes <- c("diagonal", "spherical", "full")
fits   <- lapply(shapes, function(s)
  normal_block(nb_data, blocks = selected$q, model = "mean",
               control = NB_control(verbose = FALSE, noise_covariance = s)))
data.frame(
  covariance = shapes,
  nb_param   = sapply(fits, `[[`, "nb_param"),
  loglik     = round(sapply(fits, `[[`, "loglik"), 1),
  BIC        = round(sapply(fits, `[[`, "BIC"), 1)
)
```

BIC agrees with the default here. That is not a reason to forget the full
$\Sigma$ though: the three shapes answer different questions, and a diagonal
one says nothing about how proteins co-vary once the subtype and the cluster
structure are accounted for -- which is exactly what the next section looks
at.

# Sparsifying the residual covariance

If the residual associations *are* the object of interest, the full $\Sigma$ is
required -- and asking for `sparsity > 0` selects it automatically, since a
penalty on a diagonal precision matrix would have nothing to act on. A dense
$163 \times 163$ precision matrix is unreadable though, and poorly determined
from 346 observations. Keeping the selected clustering fixed,
a graphical-lasso penalty on $\Sigma^{-1}$ addresses both at once, turning it
into a network of *conditional* associations between proteins, given the subtype
and the cluster structure.

```{r, sparse-fit}
C_selected <- model.matrix(~ 0 + factor(selected$clustering))
NB_sparse  <- normal_block(nb_data, blocks = C_selected, sparsity = 0.4,
                           model = "mean", control = NB_control(verbose = FALSE))
NB_sparse$model_par$Omega |> dim()
paste0(NB_sparse$n_edges, " edges out of ", choose(ncol(Y), 2), " possible ones.")
```

```{r, sparse-network, fig.width=6, fig.height=6}
NB_sparse$plot_network(output = "corrplot")
```

Passing `sparsity = TRUE` instead of a single value explores a whole path of
penalties and returns a collection, selected by BIC or EBIC as usual. Be aware
that each penalty triggers a graphical lasso on a $p \times p$ matrix at *every*
EM iteration: on this dataset a full path costs a couple of orders of magnitude
more than the single fit above, which is why a fixed penalty is used here.

# References
