---
title: "Analyzing multivariate Gaussian data with the Normal-Block model - First steps"
author: "Jeanne Tous, Julien Chiquet"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Analyzing multivariate Gaussian data with the Normal-Block model}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Preliminaries

This vignette illustrates the use of the normal-block function and the methods accompanying the R6 normal-block classes.

From a statistical point of view, the function normal-block adjusts a multivariate Normal-Block model (Gaussian graphical model with a latent clustering structure) to a table of observations, possibly after correcting for the effects of covariates. Depending on the parameters given, the function uses the clustering that is given or infers one, integrates zero-inflation or not, infers a sparse network or not... The parameters inference can either be done with an integrated variational Expectation-Maximization approach (recommended) or with a heuristic approach.

Theoretical explanations for this model can be found in the pre-print "An integrated method for clustering and association network inference" (Chiquet et al., 2024).

### Requirements

```{r, set-up}
library(normalblockr)
library(pheatmap)
library(paletteer)
```

# Data simulation

We illustrate the analysis with two simulated datasets.

Fix the seed to make the results reproducible.

```{r, set-seed}
set.seed(1)
```

Simulate data with generate_normal_block_data.

## Fix simulation parameters

Several parameters need to be defined to simulate the data:

```{r, fix-simulations-parameters}
n = 100 # Number of samples (number of rows in the matrix of observations)
p = 40  # Number of entities observed (number of columns in the matrix of observations)
d = 2   # Number of covariates
Q = 3   # Number of clusters
kappa = 0 # Mean zero-inflation probability (can also be a vector to define one ZI-probability for each variable). kappa = 0 means that there is no zero-inflation.
omega_structure = "erdos-renyi" # Network structure.
u_v = c(0.3, 0.1) # Parameters to generate an association matrix from a graph, details given in the bibliography.
SNR = 0.75 # Signal to Noise Ratio, defines the relative weight of the covariates and the variance
alpha = rep(1/Q, Q) # Vector giving probabilities of belonging to each cluster
range_X = c(0, 10)  # Min and max values for the covariates 
range_D = c(0.5, 1.5) # Min and max values for the individual entities variances 
```

## Simulation 1 (no zero-inflation)
**generate_normal_block_data** generates data under the Normal-Block model. It returns a list that contains the simulated covariates $X$ and observations $Y$, the simulation parameters (including the clustering $C$). 

```{r, simulation1}
my_nb_data <- generate_normal_block_data(n, p, d, Q, kappa, omega_structure, u_v,
                                        SNR, alpha, range_X, range_D)
```

```{r, simulation1-visualization}
pheatmap::pheatmap(my_nb_data$Y, 
                   color =paletteer::paletteer_c("ggthemes::Orange-Gold", n = 100), cluster_rows = FALSE, cluster_cols = FALSE, show_rownames = FALSE)
```


## Simulation 2 (with zero-inflation)

To generate zero-inflated data, we set-up kappa as a vector of zero-inflation probabilities. One needs to ensure that these values are between 0 and 1.
```{r, simulation2}
kappa_zi <- rnorm(p, mean = 0.7, sd = 0.05)
kappa_zi <- unlist(lapply(kappa_zi, f <- function(x) return(max(0, min(x, 0.9)))))
my_nb_data_zi <- generate_normal_block_data(n, p, d, q = 5, kappa_zi, omega_structure,
                                            u_v, SNR, alpha, range_X, range_D)
```
For a better visualization of the zero-inflated data, we set-up the 0 to be shown in white.

```{r, simulation2-visualization}
min_val <- min(my_nb_data_zi$Y) ; max_val <- max(my_nb_data_zi$Y)
orange_gold_pal <- paletteer::paletteer_c("ggthemes::Orange-Gold", n = 100)
n_breaks <- 100
zero_pos <- round((0 - min_val) / (max_val - min_val) * n_breaks) + 1
custom_pal <- c(orange_gold_pal[1:(zero_pos - 1)], "white", orange_gold_pal[zero_pos:n_breaks])
pheatmap::pheatmap(my_nb_data_zi$Y,
                   color = custom_pal,
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   show_rownames = FALSE)
```


# Prepare the data for a normal-block analysis
A specific data object of class **NB_data** needs to be created to analyse the data with normalblockr.

One can either use Y and X only to generate the **NB_data** object.

```{r, NB_data}
my_data    <- NB_data$new(my_nb_data$Y, my_nb_data$X)
my_data_zi <- NB_data$new(my_nb_data_zi$Y, my_nb_data_zi$X)
```

Alternatively, if one only wants to include some of the covariates contained in X, a formula can be used to specify which one:

```{r, NB_data-alt}
colnames(my_data$X) <- c("X1", "X2")
my_data_alt    <- NB_data$new(my_data$Y, my_data$X, formula = ~ 0 + X1)
```


# Run a normal-block analysis
All normal-block analyses can be run with the **normal_block** function. The function is called with different parametrisations depending on whether the clustering or the number of clusters is known or not, and according to the requested level of sparsity, among other factors. Details about this can be found in the function's documentation. 

By default, the parameters are inferred using a variational Expectation Maximization approach with 100 iterations. A finer control of the optimization is possible with the **control** parameter of **normal_block** that must be a list generated by the **NB_control** function.

## With non-zero-inflated data
### Fixed clustering
When the variables' clustering is known, one can directly give them as an input to the **normal_block** function.

```{r, simulation1-NB1}
my_NB <- normal_block(data = my_data,
                      blocks = my_nb_data$parameters$C)
```

The model convergence can be checked with **plot**.

```{r, simulation1-NB1-plot}
plot(my_NB)
```

A summary of the results is accessible via **print**.
```{r, simulation1-NB1-print}
print(my_NB)
```
The inter-clusters association network can be visualized with the **plot_network** function.

```{r, simulation1-NB1-plot_network}
 my_NB$plot_network()
```


### Fixed number of clusters
When the variables' clustering is unknown, one can simply fix the number of clusters in the **blocks** parameter.

```{r, simulation1-NB2}
my_NB <- normal_block(data = my_data,
                      blocks = 3)
print(my_NB)
plot(my_NB)
```
The clustering inferred by the **normal_block** function can be compared with the real clustering if it is known.

```{r, simulation1-ARI}
aricode::ARI(my_NB$clustering, apply(my_nb_data$parameters$C, 1, which.max))
```


### Unknown number of clusters
When the number of clusters is unknown, one can give the **normal_block** function a range of possible numbers and get a collection of Normal-Block models, one per number of clusters.

```{r, simulation1-NB3}
my_NB_unknown <- normal_block(data = my_data,
                      blocks = 2:5)
```
**my_NB_unknown** is a collection of Normal-Block models. It is possible to select a specific one either by picking a number of clusters or by selecting the best model for a given criterion (BIC, deviance, EBIC, ICL). The values for these criteria for each model can be seen with the **plot** function.

```{r, simulation1-NB3-plot}
plot(my_NB_unknown)
```

The model selection is done with **get_model** when a specific number of clusters is required and with **get_best_model** to select the best model for a given criterion.

```{r, simulation1-NB3-model-selection}
myNB_3   <- my_NB_unknown$get_model(3)
myNB_BIC <- my_NB_unknown$get_best_model("BIC")
```


### Playing with sparsity
By default, no penalty is applied on the association network. One can add a penalty by using the sparsity parameter of the **normal_block** function. This can be done with any of the above parametrisation (fixed clustering, fixed number of clusters, unknown number of clusters). Let us show an example with a fixed clustering.

If one wants to use a fixed $l_1$ penalty (see Tous & Chiquet, 2024) on the network, that value should be given in the sparsity parameter.
```{r, simulation1-NB-fixed-sparsity}
my_NB_sparse_low <- normal_block(data = my_data,
                                 blocks = my_nb_data$parameters$C,
                                 sparsity = 0.1)
my_NB_sparse_high <- normal_block(data = my_data,
                                  blocks = my_nb_data$parameters$C,
                                  sparsity = 10)
```
The bigger the penalty, the sparser the network.
```{r, simulation1-NB-low-sparsity-plot}
my_NB_sparse_low$plot_network()
```
```{r, simulation1-NB-high-sparsity-plot}
my_NB_sparse_high$plot_network()
```
It is usually hard to know a priori what would be the best sparsity penalty. One can allow the **normal_block** function to explore different sparsity levels by simply setting **sparsity = TRUE**. The **normal_block** function then creates a collection of Normal-Block models, one per sparsity penalty.

```{r, simulation1-NB-changing-sparsity}
my_NB_sparse <- normal_block(data = my_data,
                             blocks = my_nb_data$parameters$C,
                             sparsity = TRUE)
```

The different criteria can then be plotted as a function of the penalty, using **plot**.
```{r, simulation1-NB-changing-sparsity-plot}
plot(my_NB_sparse)
```
Penalty selection can be done similarly to the selection of the number of clusters.
```{r, simulation1-NB-changing-sparsity-model-selection}
myNB_sparse_0.1   <- my_NB_sparse$get_model(0.1)
myNB_sparse_BIC   <- my_NB_sparse$get_best_model("BIC")
```

One can also make both the sparsity penalty and the number of clusters vary. 

```{r, simulation1-NB-sparse_unknown}
my_NB_sparse_unknown <-  normal_block(data = my_data,
                                      blocks = 2:6,
                                      sparsity = TRUE)
```

The result is a collection of normal-block models with different numbers of clusters and ifferent penalties.
```{r, simulation1-NB-sparse_unknown-who-am-I}
my_NB_sparse_unknown$who_am_I
```


The plot function allows one to analyse each criterion as a function of the number of clusters and of the penalties. By default, the penalties to be tested are computed by the **normal_block** function for each number of clusters and they can be different from one number to another, this explains the blanks in the plot.
```{r, simulation1-NB-sparse_unknown-plot}
plot(my_NB_sparse_unknown, "BIC")
```

Model selection is also done using the **get_model** and **get_best_model** functions. Using  **get_model**, one can either fix the n umber of clusters only and get a collection of model with different sparsity levels, or fix both the number of clusters and the sparsity level. 

```{r, simulation1-NB-sparse_unknown-selection}
my_NB_sparse_3     <- my_NB_sparse_unknown$get_model(3)
my_NB_sparse_3_0.1 <- my_NB_sparse_unknown$get_model(3, 0.1)
```

## With zero-inflated data
The process is similar with zero-inflated data but one needs to specifgy that the data is zero-inflated in the **normal_block** function. We show an example with a fixed number of blocks and no penalty on the network.

```{r, simulation2-NB}
my_NB_zi <- normal_block(data = my_data_zi,
                         blocks = 4,
                         zero_inflation = TRUE)
```
When the data is zero-inflated, the inference process may take longer.

```{r, simulation2-NB-plot}
plot(my_NB_zi)
```


The clustering inference may also be harder.
```{r, simulation2-ARI}
aricode::ARI(my_NB_zi$clustering,
             apply(my_nb_data_zi$parameters$C, 1, which.max))
```



