---
title: "Initializing the unknown clustering: which heuristic, and why"
format: html
editor: visual
---

## Motivation

`NormalBlockUnknownClusters` (and the collections built on it) need a
starting clustering of the `p` variables before the (V)EM recursion can run.
This starting point comes from `private$clustering_methods` in
`R/NormalBlockBase.R`, selectable via `NB_control(clustering_approx = ...)`.

The question behind this whole notebook: **the normal-block model's
clustering target is the *covariance* structure of the residuals, not their
*values*** -- for two variables $j, j'$ in the same cluster $k$,
$\mathrm{Cov}(Y_j, Y_{j'}) = \mathrm{Var}(W_k)$, a constant shared by every
pair in the cluster, regardless of each variable's own idiosyncratic noise
$D_j$. A heuristic that clusters on raw residual proximity (plain `kmeans`)
is targeting something else entirely. So: does a heuristic that is more
faithful to this covariance structure actually initialize better, and is
there a single best choice?

Short answer, confirmed below on two real datasets (breast cancer
proteomics, university webpages text data): **no single heuristic dominates
across datasets.** The package therefore keeps a menu of 5, rather than
picking one "best" method.

## Setup

```{r useful-libraries, message=FALSE}
library(normalblockr)
library(dplyr)
library(ggplot2)
```

```{r data-breast-cancer}
base_dir <- "../1_breast_cancer_proteomics"
RPPA_expr <- readRDS(file.path(base_dir, "RPPA_expr.rds"))
clinical  <- readRDS(file.path(base_dir, "brca_clinical.rds"))
RPPA_clin <- clinical[match(rownames(RPPA_expr), clinical$sampleId), ]
data_brca <- NormalBlockData$new(
  scale(as.matrix(RPPA_expr)),
  model.matrix(~ 0 + PAM50_SUBTYPE, data = RPPA_clin)
)
```

```{r data-university-webpages, eval=FALSE}
## see ../2_university_webpages/normal_block_analysis_university_webpages_data.qmd
## for the full text-preprocessing pipeline (tm-based DTM construction,
## entropy-based term selection, log(1+freq) + scale()); reproduced here only
## as a pointer, not re-run, since it requires the `tm` package and the raw
## corpus.
data_web <- NormalBlockData$new(Y_web, X = matrix(1, nrow(Y_web), 1))
```

## The 5-method menu

```{r show-registry, eval=FALSE}
## R/NormalBlockBase.R, private$clustering_methods
list(
  kmeans    = function(R, q) kmeans(t(R), q, nstart = 30, iter.max = 50)$cluster,
  ward2     = function(R, q) cutree(hclust(dist(1 - cor(R)), method = "ward.D2"), q),
  sbm       = function(R, q) { # via sbm_clustering_path() for collections over several q
    options <- list(verbosity = 0, exploreMin = q, exploreMax = q, plot = FALSE, nbCores = 1)
    mySBM <- sbm::estimateSimpleSBM(cov(R), "gaussian", estimOptions = options)
    mySBM$setModel(q)
    mySBM$memberships
  },
  kmeansvar = function(R, q) ClustOfVar::kmeansvar(X.quanti = R, init = q, nstart = 30)$cluster,
  spectral  = function(R, q) {
    U <- eigen(cov(R), symmetric = TRUE)$vectors[, seq_len(q), drop = FALSE]
    U <- U / pmax(sqrt(rowSums(U^2)), 1e-10)
    kmeans(U, q, nstart = 30, iter.max = 50)$cluster
  }
)
```

| method | targets | theoretical fit to the model | cost |
|---|---|---|---|
| `kmeans` | residual *values* (Euclidean) | weak -- not the model's actual clustering criterion | cheap |
| `ward2` | pairwise *correlation* | partial -- confounded by heterogeneous $D_j$ (correlation divides by $\sqrt{\mathrm{Var}(W_k)+D_j}$) | cheap |
| `sbm` | block-constant *covariance* (Gaussian SBM) | exact match -- a Gaussian SBM literally assumes constant within/between-block covariance | expensive (mitigated for collections by `sbm_clustering_path()`, see "Implementation notes" in `inst/normal_block_models.qmd`) |
| `kmeansvar` | one dominant factor per cluster, *arbitrary* loadings | related but more permissive than the model (which assumes a *unit* loading) | expensive |
| `spectral` | top-$q$ eigenvectors of $\mathrm{cov}(R)$, row-normalized | same target as `sbm`, via a cheap proxy (classical for SBM recovery) instead of a combinatorial search | cheap |

## Benchmark protocol

Both datasets, $q = 2$ to $25$. For each $(q, \text{method})$: get the
heuristic clustering on the OLS residuals (`ols_residuals()`,
`R/utils.R`), seed `NormalBlockUnknownClusters$new(data, q, control =
NB_control(clustering_init = cl))`, run `$optimize()`, record the final
ICL. **Strictly sequential** -- a single R process, no `parallel`/
`mclapply`/`foreach` (the BLAS backend here is already multithreaded across
all cores; comparisons must not also be parallelized on top of that, or
timings become meaningless).

```{r benchmark-loop, eval=FALSE}
q_values <- 2:25
methods  <- c("kmeans", "ward2", "kmeansvar", "sbm", "spectral")

run_one <- function(data, q, method) {
  R <- ols_residuals(data)
  cl <- if (method == "sbm") {
    sbm_clustering_path(R, q_values)[[as.character(q)]] # one shared exploration
  } else {
    dummy <- NormalBlockUnknownClusters$new(data, q, control = NB_control(verbose = FALSE))
    dummy$.__enclos_env__$private$clustering_methods[[method]](R, q)
  }
  model <- NormalBlockUnknownClusters$new(data, q, control = NB_control(verbose = FALSE, clustering_init = cl))
  model$optimize(control = NB_control(verbose = FALSE))
  data.frame(q = q, method = method, niter = length(model$objective), ICL = model$ICL)
}

results <- do.call(rbind, lapply(methods, function(m) do.call(rbind, lapply(q_values, run_one, data = data_brca, method = m))))
```

## Results

```{r load-results}
menu <- readRDS("results/menu_benchmark.rds")
```

```{r icl-vs-q-plot, fig.width=9, fig.height=4}
ggplot(menu, aes(q, ICL, color = method)) +
  geom_line() + geom_point(size = 0.8) +
  facet_wrap(~ dataset, scales = "free_y") +
  theme_minimal() +
  labs(title = "ICL vs number of clusters, by initialization method")
```

```{r rank-table}
menu %>%
  group_by(dataset, q) %>%
  mutate(rank = rank(ICL, ties.method = "average")) %>%
  ungroup() %>%
  group_by(dataset, method) %>%
  summarise(mean_rank = round(mean(rank), 2), wins = sum(rank == 1), .groups = "drop") %>%
  arrange(dataset, mean_rank) %>%
  knitr::kable()
```

**No method wins on both datasets.** `sbm` is best on breast cancer
proteomics (9/24 wins) but worst on university webpages (0/24); `kmeansvar`
is the mirror image (worst on breast cancer, co-best on webpages);
`spectral` is mediocre on breast cancer but co-best on webpages, at a
fraction of `kmeansvar`'s cost. `kmeans` is never the *worst* on either
dataset, which is why it was chosen as the default in `NB_control()` -- it
optimizes for "never embarrassing" rather than "sometimes best".

## Side investigation 1: a hand-rolled Armadillo k-means

Tested whether reimplementing k-means in RcppArmadillo (Lloyd's algorithm,
multi-start) could beat `stats::kmeans()` (Hartigan-Wong) on speed and/or
quality. Reference implementation kept at `results/kmeans_arma.cpp` (not
compiled into the package -- exploratory only, via `Rcpp::sourceCpp()`).

```{r kmeans-arma-results}
ka <- readRDS("results/kmeans_arma_benchmark.rds")
ka %>%
  group_by(dataset, method) %>%
  summarise(total_cluster_time = round(sum(cluster_time), 2), .groups = "drop") %>%
  knitr::kable(caption = "Total clustering time (s), summed over q in {3,8,15,22} x nstart in {10,30,100,300}")
```

Findings:

- **Speed**: no systematic win -- faster on one dataset, slower on the other.
- **Quality (WCSS)**: plain Lloyd's is *consistently worse* than
  Hartigan-Wong (28/32 configurations), by ~2.6% on average, and this gap
  does **not** close with more restarts: `kmeans_arma` at `nstart=300` still
  doesn't beat `stats::kmeans` at `nstart=10` in most cases. The gap is the
  local-search algorithm itself, not an insufficient number of restarts.

```{r kmeanspp-results}
kapp <- readRDS("results/kmeans_arma_pp_benchmark.rds")
kapp %>%
  group_by(method, nstart) %>%
  summarise(total_cluster_time = round(sum(cluster_time), 3), .groups = "drop") %>%
  knitr::kable(caption = "Total clustering time (s), summed over q in {3,8,15,22}, both datasets")
```

Following up: does k-means++ seeding let `kmeans_arma` match `stats::kmeans`
with far fewer restarts? Confirmed on the resource-efficiency axis -- WCSS
plateaus by `nstart = 3-5` instead of needing hundreds -- but the plateau
itself is still *worse* than Hartigan-Wong's. **Conclusion: not worth
shipping.** No clear edge on either axis once Hartigan-Wong is the
baseline; the existing 5-method menu already provides cheap diversity.

## Side investigation 2: which spectral-clustering variant matters

Classical spectral clustering has several normalization conventions
(unnormalized adjacency, symmetric-normalized Laplacian, row-normalized
embedding, ...). Tested 6 combinations: eigenvectors of $\mathrm{cov}(R)$ vs
$\mathrm{cor}(R)$ (= the symmetric-Laplacian analogue, since $\mathrm{cor} =
D^{-1/2} \Sigma D^{-1/2}$ with $D = \mathrm{diag}(\Sigma)$), each with/without
row-normalizing the eigenvector embedding (Ng-Jordan-Weiss), plus a
diagonal-deleted variant motivated by the same $D_j$-interference argument
from the motivation section: $\mathrm{Cov}(Y_j, Y_{j'})$ for $j \ne j'$ never
involves $D_j$, only the *diagonal* of $\mathrm{cov}(R)$ does, so a few
high-noise variables can hijack top eigenvector slots if the diagonal is
left in.

```{r spectral-variants-synthetic, eval=FALSE}
## synthetic stress test: heteroscedastic noise, not aligned with cluster
## membership -- 30 replications, p=60, q=4, n=80
## mean ARI with diagonal kept:   0.0025  (clustering destroyed)
## mean ARI with diagonal zeroed: 0.342   (real signal recovered)
## nodiag better in 30/30 replications
```

That diagonal-deletion fix is decisive in this adversarial synthetic
setting. On the two **real** datasets, though, it is not what matters most:

```{r spectral-variants-results}
sv <- readRDS("results/spectral_variants_benchmark.rds")
sv %>%
  group_by(dataset, q) %>%
  mutate(rank = rank(ICL, ties.method = "average")) %>%
  ungroup() %>%
  group_by(dataset, method) %>%
  summarise(mean_rank = round(mean(rank), 2), wins = sum(rank == 1), .groups = "drop") %>%
  arrange(dataset, mean_rank) %>%
  knitr::kable()
```

Row-normalization (`*_rownorm`) dominates the other 4 variants on both
datasets; diagonal-deletion does not show the same benefit here, probably
because the real noise heterogeneity isn't as extreme as the synthetic
stress test, and row-normalization already absorbs much of the same
"a few extreme embedding rows distort kmeans" failure mode. **This is why
`spectral` in the registry is specifically `cov(R)` eigenvectors with
row-normalization** -- the other variants were measurably worse and didn't
make the cut.

## Open threads for later

- The diagonal-deletion idea was validated synthetically but not on real
  data here -- worth retesting on a dataset with more extreme,
  cluster-misaligned noise heterogeneity than these two happen to have.
- Only two datasets so far. A third (e.g. `3_MICROCOSM`) would help tell
  whether `kmeans`'s "never worst" property generalizes or was a
  two-dataset coincidence.
- `kmeansvar` and `spectral` were close to tied on university webpages
  (12/24 wins each) -- worth a paired significance check rather than raw
  win-counting if this matters for a future default-choice decision.
