---
title: "Breast cancer proteomics data analysis [Brigham & Women's Hospital & Harvard Medical School (2012)]"
format: html
---

## Useful libraries

```{r useful-libraries, message=FALSE}
library(tidyverse)
library(normalblockr)
library(aricode)
library(clusterProfiler)
library(pathview)
library(enrichplot)
library(org.Hs.eg.db)
set.seed(1)
```

## Loading & formatting data

Loading proteomics data to analyse and clinical data to get cancer subtypes as covariates.

```{r data-load}
data(brca_rppa)
```

```{r Normal-Block-formatting}
Y          <- as.matrix(brca_rppa$expr)
X_subtype <- model.matrix(~ 0 + PAM50_SUBTYPE, data = brca_rppa$covariates)
data_subtype <- NormalBlockData$new(Y, X_subtype)
```

## Applying Normal-Block with a fixed, data-driven grouping

Before letting Normal-Block infer its own grouping of the proteins, we build a
simple, fully data-driven baseline: a hierarchical (Ward) clustering of the
proteins from their expression profiles alone, with no reference to the
Normal-Block model. By default `normalblockr` scales the data, so the a
priori clustering is computed on the same (column-)scaled matrix for
consistency. Six clusters is an arbitrary but visually natural cut of the
dendrogram (highlighted below); the unsupervised exploration in the next
section will pick its own number of blocks independently.

```{r hclust-based group}
hc_expr <-
  brca_rppa$expr |> scale() |> t() |> dist() |> hclust("ward.D2")
plot(hc_expr, labels = FALSE, hang = -1,
     main = "Hierarchical clustering of proteins (Ward, on scaled expression)",
     xlab = "proteins", sub = "")
rect.hclust(hc_expr, k = 6, border = "red")
group <- cutree(hc_expr, 6) |> normalblockr:::as_indicator()
```

This fixed grouping is then handed to Normal-Block as known clusters -- the
model only estimates the association network between the 6 blocks, not the
grouping itself.

```{r running-normal-block-known-group}
NB_prot_group <- normal_block(data_subtype, blocks = group)
plot(NB_prot_group)
```

## Applying Normal-Block with an unknown, inferred grouping

### Running the integrated inference procedure for 1 to 30 clusters.

```{r running-normal-block}
NB_prot_subtype <- normal_block(data_subtype, blocks = 1:30)
```

### Model selection: criteria to fix the number of clusters

```{r plotting criteria}
NB_prot_subtype$plot(c("deviance", "BIC", "ICL"))
```

### Refinement: try to enhance optimisation based on ICL shape

```{r refine}
NB_prot_subtype$refine()
NB_prot_subtype$plot(c("deviance", "BIC", "ICL"))
```

### Select the best model reached

```{r model-selection}
selected_NB <- NB_prot_subtype$get_best_model("ICL")
paste0("ICL selects ", selected_NB$q, " clusters.")
```

## Enrichment analysis

Phospho-site antibodies (e.g. `EGFR_PY1068`) carry a placeholder, *negative*
`entrezGeneId` (there is no separate gene for "EGFR phosphorylated at
Y1068", only EGFR itself -- same convention already handled in
`data-raw/brca_rppa_go_annotation.R`). Left as-is, `enrichKEGG` cannot map
those IDs to anything, which both throws away ~a quarter of the proteins
and triggers clusterProfiler's "no gene can be mapped" warning for any
cluster unlucky enough to be made up mostly of phospho-sites. Resolving
each phospho-site to its underlying gene's real Entrez ID via its symbol
fixes both.

```{r enrichment}
ga         <- brca_rppa$gene_annotation
is_phospho <- ga$type == "phosphoprotein"
gene_symbol      <- sub("_.*", "", ga$protein)
symbol_to_entrez <- AnnotationDbi::select(org.Hs.eg.db, keys = unique(gene_symbol[is_phospho]),
                                           columns = "ENTREZID", keytype = "SYMBOL")
EntrezGeneID <- ga$entrezGeneId
EntrezGeneID[is_phospho] <- symbol_to_entrez$ENTREZID[match(gene_symbol[is_phospho], symbol_to_entrez$SYMBOL)]
EntrezGeneID <- setNames(EntrezGeneID, ga$protein)

df_inter <- data.frame(Entrez = EntrezGeneID, group = selected_NB$clustering)
inter_kegg <- compareCluster(Entrez~group, data = df_inter, fun='enrichKEGG', pvalueCutoff = 0.001)
inter_kegg <- setReadable(inter_kegg, OrgDb = "org.Hs.eg.db", keyType="ENTREZID")
dotplot(inter_kegg, font.size = 8)
```

## Comparison with the data-driven clustering

How much does Normal-Block's own, unsupervised grouping of the proteins
agree with the purely data-driven hierarchical clustering used as a fixed
baseline above? For every q explored in `NB_prot_subtype`, we compute the
Adjusted Rand Index (ARI) between its inferred clustering and the q = 6
hierarchical-clustering baseline (blue dashed line); the red dashed line
marks the model ICL actually selected.

```{r comparison-with-hclust}
df_ari <- tibble(
  q   = purrr::map_dbl(NB_prot_subtype$models, "q"),
  ARI = purrr::map_dbl(purrr::map(NB_prot_subtype$models, "clustering"),
                        aricode::ARI, NB_prot_group$clustering)
)
ggplot(df_ari, aes(x = q, y = ARI)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = ncol(group), linetype = "dashed", colour = "blue", alpha = .5) +
  geom_vline(xintercept = selected_NB$q, linetype = "dashed", colour = "red", alpha = .5) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  ggtitle(label = "Agreement with the hierarchical-clustering baseline",
          subtitle = "blue: baseline's q -- red: ICL-selected q") +
  xlab("number of clusters (q)") + ylab("Adjusted Rand Index") +
  theme_bw()
```
